---
title: "Plumbing R Code for the web"
author: "Remko Duursma"
date: 2018-06-06
categories: ["R"]
tags: ["web", "api", "plumber", "optparse"]
description: "My experience with the plumber package to deploy a REST service in R."
---


```{r setup, include=FALSE, cache=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE,
               eval = FALSE, 
               out.width="600px", 
               dpi=400, 
               warning = FALSE)
```


I recently had the pleasure to complete an assessment - as part of a job interview. The assessment was enjoyable exactly because I did not have much - if any - experience with what I was supposed to be assessed on. But, at least for me, that's the point: how quick can we learn new tools in the data science toolkit?

Unfortunately I am not going to share the actual assessment (or even some part of it), but will talk about the `plumber` package which I used to deploy an API.

The goal is to end up with a 'REST' service, which allows you to browse to a URL like this,

```
<server ip address>:8001/api/items/nr123/predict
```

where the IP address of the server is used, or 127.0.0.1 if you are running a 'local host'. Here we communicate via port 8001 - but this could be any open port on your system/server. Your browser will then return some text output (perhaps : "prediction: 15", if the service predicts the remaining lifetime of some item).

The `plumber` package can be used to quickly and effectively deploy this kind of API, all from the comfort of your R code. 


## Dynamic routes

I am not going to repeat any of the introductory stuff of [the excellent plumber documentation](), just highlight some key points. We first make a file with the API definitions, using some special formatting. Suppose the file 'item_predictor_api_definition.R' contains the following:


```{r}
#* Predict item lifetime
#* @param id
#* @get /prediction
function(item){
  
  my_prediction_function(item)
  
}
```

Here we define a single API endpoint - but not quite in the format that we wish to have, because the request to this endpoint will have to look like this:

```
<server ip address>:8001/prediction?item="nr123"
```

I suppose this is OK too, but a common recommendation is to allow a URL like `/api/item/nr123/prediction`. The key `nr123` (referring to some item ID) is therefore the dynamic bit of the URL. We can easily tell `plumber` what to do:

```{r}
#* Predict item lifetime
#* @param id
#* @get /api/item/<item>/prediction
function(item){
  
  my_prediction_function(item)

}
  
```

Note how the Roxygen field calls the parameter `id`, not `item` as you would expect. I found it necessary in plumber version 0.4.5 to **not** use the same name as the function argument there - otherwise the service crashes with a weird message ([see this issue I openened](https://github.com/trestletech/plumber/issues/267)).


## Deploy time!

With our API endpoint defined (of course we can add as many as we like), we can start the API via:

```{r}
library(plumber)

# There are no other arguments besides a directory to find the file.
p <- plumb('item_predictor_api_definition.R')

# Deploy!
p$run(port=8001)
```

Your R process will now be busy (until you close it), with a message like:

```
Starting server to listen on port 8001
```

And so our work is done. Next, it makes sense to save the API deployment code in a script, let's say it is called `item_predictor_deploy_api.R`. Then we can simply deploy our API from the command line with:

```
Rscript item_predictor_deploy_api.R
```


## Supplying command line arguments


Better yet, it would be neat to pass the port to open up as an argument in the command line. To do this, I used the `optparse` package which is a bit verbose in its usage, but super effective.

All we have to do is preface the API deploy script with some `optparse` code. We also want to make sure the user is told off when no port is defined (a potential security hazard).


```{r}
library(optparse)

# Parse command line arguments and display help.
option_list = list(
  make_option(c("-p", "--port"), type="integer", default=NULL, 
              help="Port number for REST API", metavar = "integer")
)

opt_parser <- OptionParser(option_list=option_list,
                           description="Deploy API for item predictor.")
opt <- parse_args(opt_parser)


# If the user does not supply the port - 
# fail with an error and a message, and display the help usage.
if(is.null(opt$port)){
  print_help(opt_parser)
  stop("Must supply port to open for REST API.",
       call. = FALSE)
}


# Now continue with the API.
library(plumber)

# Set the plumbing for the API.
p <- plumb("R/api_definitions.R")

# Deploy.
# Note opt$port is the value read from the command line argument
p$run(port = opt$port)
```


This is pretty neat because now we can do from the command line:

```
Rscript item_predictor_deploy_api.R --port 8001
```

This makes it very easy for your system admin to deploy this particular API on any port they wish, instead of needing to poke around in our code.

Also neat is what happens when the user does not define a port:

```
Usage: item_predictor_deploy_api.R [options]
Deploy API for item predictor.

Options:
        -p INTEGER, --port=INTEGER
                Port number for REST API

        -h, --help
                Show this help message and exit


Error: Must supply port to open for REST API.
Execution halted
```

We get not just an error message, but also a reminder how the script should be used. And all of this is outside R - this is a massive advantage when you are deploying a service for R-agnostics.


## Conclusions

Obviously this was no introduction to `plumber`, just a note on some settings I collected during my use of the package. The neat organization of the API definitions, and the code to deploy the API, together with `optparse` for command line options makes for an easy to use service to deploy a REST API. 

For larger scale services, we of course wonder what happens if our single process we opened is busy - and another user sends a request. The `plumber` folks describe [how Docker can be used for maintaining multiple processes, load balancing, etc.](https://www.rplumber.io/docs/hosting.html#hosting) Certainly something to try for next time.

































